Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-08 text/ms.tex
Clean up some comments
Small edits

% TODO: hyphenate the adjective 'black swan'? E.g. black-swan event, black swans.
\section{Abstract}

@@

There is compelling anecdotal evidence for ecological black swans, but systematic evidence has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}. In fact, anecdotes of population catastrophes are numerous and catastrophes may be the most important element affecting population persistence \citep{mangel1994}. For marine mammal populations, we have compelling evidence of catastrophes \citep{gerber2001, ward2007}, and recently, heavy-tailed distributions have been discovered in time-series of marine microbe abundance \citep{segura2013} and time to extinction for experimental waterflea populations \citep{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which uncovered little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.
% TODO: pull anecdote from that Nunez study

% TODO \citep{mangel1994} catastrophes may be the most important element affecting species persistence times and more common than we think
% TODO \citep{gerber2001} and \citep{ward2007} evidence for marine mammal populations

@@

Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series. We apply our model to \NPops\ populations and estimate the frequency and strength of black-swan events in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
%We find a high probability of ecological
%black swans across a wide range of taxa but in a limited number of populations.
%When they occur, extreme climate, predation, parasites, and their interactions
%are common documented causes. We find little evidence that intrinsic
%life-history characteristics are related to the probability of black swans.

@@

Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population. 
%These exemplar populations were chosen haphazardly with the purpose of generating hypotheses about the causes of population dynamic black swans.

@@

%TODO IMPORTANT: you need to quantify across all identified black swan events how many were downward and how many upward. The assumption of your model is a symmetric distribution, but if in fact downward events are more common, perhaps a skewed distribution would be more realistic. Furthermore, if there is skew this would *increase* the probability of higher extinction risk. Something to definitely highlight in the results and abstract.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? 
In recent decades ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.

@@

We thank Justin D. Yeakel, TODO, and other members of the Earth to Ocean research group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Three silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).

>>>>

\section{Abstract}

@@

There is compelling anecdotal evidence for ecological black swans, but systematic evidence across taxa has been elusive. A survey of ecologists indicated that surprising outcomes of field experiments are far more common than we assume \citep{doak2008}, and events such as the global invasion of  Argentine ants and the mutation of viruses to infect new hosts could be considered black swans \citep{nunez2012}. In fact, anecdotes of population catastrophes are numerous and catastrophes may be the most important element affecting population persistence \citep{mangel1994}. For marine mammal populations, we have compelling evidence of catastrophes \citep{gerber2001, ward2007}, and recently, heavy-tailed distributions have been discovered in time-series of marine microbe abundance \citep{segura2013} and time to extinction for experimental waterflea populations \citep{drake2014}. As far as we can tell there have been two systematic surveys for ecological black swans: one with North American breeding birds and the other with the Global Population Dynamics Database, which uncovered little clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological challenges to their detection \citep{allen2001,ward2007}.

@@

Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series. We apply our model to \NPops\ populations and estimate the frequency and strength of black-swan events in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.

@@

Finally, we investigated a sample of populations that our method categorized as having a high probability of heavy tails (Pr$(\nu < 10) > 0.5$). Where possible, we found the documented causes of ecological black swans in the primary data source cited in the GPDD or in other literature describing the population.

@@

%TODO IMPORTANT quantify across all identified black swan events how many were downward and how many upward. The assumption of your model is a symmetric distribution, but if in fact downward events are more common, perhaps a skewed distribution would be more realistic. Furthermore, if there is skew this would *increase* the probability of higher extinction risk. Something to definitely highlight in the results and abstract.

@@

Most importantly, what is the impact of allowing for black swans in forecasts of ecological risk? 
In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes \citep{meehl2004,ipcc2012}.
Financial analysts are concerned with the shape of financial-return-distribution tails because these impact estimates of risk --- the probability of a specific magnitude of undesired event occurring \citep{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events.

@@

We thank Justin D. Yeakel, TODO, and other members of the Earth to Ocean research group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Silhouette images were obtained from \texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported license; sources are listed in the Supporting Information. Funding was provided by a Simon Fraser University Graduate Fellowship (SCA), the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the Canada Research Chairs Program (NKD).
